---
title: "Manipulating observations (rows) with `dplyr`"
teaching: 50
exercises: 30
questions:
- "How to order rows in a table?"
- "How to retain only unique rows in a table (no duplicates)?"
- "How to choose rows of a table based on certain conditions?"
- "How to use conditions for a range of applications?"
- "How to save data into a file?"
objectives: 
- "Understand how logical comparisons are made and the logical data type."
- "Remember and distinguish between different types of logical operators."
- "Use conditional operations to obtain specific observations from data using the `filter()` function."
- "Apply conditional operations to clean data mistakes and/or highlight elements of interest in graphs using the `ifelse()` function."
- "Save data frame into a file."
keypoints:
- "Order rows in a table using `arrange()`. Use the `desc()` function to sort in descending order."
- "Retain unique rows in a table using `distinct()`."
- "Choose rows based on conditions using `filter()`."
- "Conditions can be set using several operators: `>`, `>=`, `<`, `<=`, `==`, `!=`, `%in%`."
- "Conditions can be combined using `&` and `|`."
- "The function `is.na()` can be used to identify missing values. It can be negated `!is.na()` to find non-missing values."
- "Use the `ifelse()` function to define two different outcomes of a condition."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("05-")
knitr::opts_chunk$set(fig.width=12, fig.height=7) 
```

```{r, include=FALSE}
# set ggplot2 theme with bigger text size
library(tidyverse)
theme_set(theme_grey() + theme(text = element_text(size = 16)))
```


In this lesson we're going to learn how to use the `dplyr` package to manipulate rows 
of our data. 

As usual when starting an analysis on a new script, let's start by loading the 
packages and reading the data. We will continue with gapminder data from 1960 to 2010:

```{r, message=FALSE}
library(tidyverse)

# Read the data, specifying how missing values are encoded
gapminder1960to2010 <- read_csv("data/gapminder1960to2010_socioeconomic.csv", 
                                na = "")
```


## Ordering Rows

Or order the rows of a table based on certain variables, we can use the `arrange()` 
function:

```{r}
gapminder1960to2010 %>% 
  select(country, world_region, year) %>% 
  # order by year
  arrange(year)
```

In this case, the rows are sorted by _increasing numeric order_ of the variable 
year. 

When ordering by a character variable, the order will be determined alphabetically:

```{r}
gapminder1960to2010 %>% 
  select(country, world_region, year) %>% 
  # order by world_region
  arrange(world_region)
```

To arrange based on _descending_ order, you can wrap the variable in the `desc()`
function:

```{r}
gapminder1960to2010 %>% 
  select(country, world_region, year) %>% 
  # order by year
  arrange(desc(year))
```

Finally, you can include several variables within `arrange()`, which will sort the 
table by each of them. For example:

```{r}
gapminder1960to2010 %>% 
  select(country, world_region, year) %>% 
  # order by year, then by world region (z-a), then country (z-a)
  arrange(year, desc(world_region), desc(country))
```


## Retain Unique Rows

Sometimes it is useful to retain rows with unique combinations of some of our variables
(i.e. remove any duplicated rows). This can be done with the `distinct()` function. 

```{r}
# get unique combination of main religion and world region
gapminder1960to2010 %>% 
  distinct(main_religion, world_region)
```


## Choosing Rows Based on Conditions

To choose rows based on specific criteria, we can use `filter()`. For example, 
to get all the rows of data from the 2000's:

```{r}
gapminder1960to2010 %>% 
  select(country, year) %>% 
  filter(year > 2000)
```


### Conditional Operations

It is important to understand that when we set a condition like above, the output 
is a _logical vector_. Let's see an example using a small vector.

```{r}
some_years <- c(1985, 1990, 1999, 1995, 2010, 2000)
some_years < 2000
```

So what the `filter()` function does is evaluate the condition and return the rows 
for which that condition is true. 

It is possible to combine several conditions by using the _logical operators_ 
`&` (AND) and `|` (OR). For example, if we wanted the years between 1990 and 2000:

```{r}
# both conditions have to be true
some_years > 1990 & some_years < 2000
```

And if we wanted the years below 1990 or above 2000, then:

```{r}
# only one or the other of the conditions has to be true
some_years < 1990 | some_years > 2000
```


> ## Conditional  operators
> 
> To set filtering conditions, use the following _relational operators_:
> 
> - `>` is greater than
> - `>=` is greater than or equal to
> - `<` is less than
> - `<=` is less than or equal to
> - `==` is equal to
> - `!=` is different from
> - `%in%` is contained in
> 
> To combine conditions, use the following _logical operators_:
> 
> - `&` AND
> - `|` OR
> 
> Some functions return logical results and can be used in filtering operations:
> 
> - `is.na(x)` returns _TRUE_ if a value in _x_ is missing
> 
> The `!` can be used to negate a logical condition:
> 
> - `!is.na(x)` returns _TRUE_ if a value in _x_ is NOT missing
> - `!(x %in% y)` returns _TRUE_ if a value in _x_ is NOT present in _y_
{: .callout}


Let's see an example on how we can combine the `filter()` function with plotting. 

In Hans Rosling's TED Talk, he compared the change in `income_per_person` and `child_mortality` 
between three countries: Uganda, Brazil and South Korea. We can filter our dataset 
to retain only these countries by using the `%in%` operator. 

```{r, eval=FALSE}
gapminder1960to2010 %>% 
  filter(country %in% c("Uganda", "Brazil", "South Korea"))
```

We will pipe this filter to ggplot and use `geom_path()` to look at the change of these 
countries over time. 
This geometry is similar to `geom_line()`, but it connects points in the order they appear 
on the table (which we can sort beforehand with `arrange()`), whereas `geom_line()` connects 
the points based on the order they appear along the x and y axis. 
(Try changing the code below to see the difference)

```{r}
gapminder1960to2010 %>% 
  filter(country %in% c("Uganda", "Brazil", "South Korea")) %>% 
  # ensure data is in order of country and year for geom_path
  arrange(country, year) %>% 
  ggplot(aes(income_per_person, child_mortality)) +
  geom_path(aes(colour = country), arrow = arrow())
```


> ## Exercise
> 
> 1. Fix the following code (where the word "FIXME" appears), to produce the plot shown below. 
> 
> ```{r, eval=FALSE}
> gapminder1960to2010 %>% 
>   mutate(population_total = FIXME) %>% 
>   filter(FIXME) %>% 
>   ggplot(aes(children_per_woman, life_expectancy)) +
>   geom_point(aes(size = population_total, colour = world_region)) +
>   facet_grid(cols = FIXME) +
>   scale_colour_brewer(palette = "Dark2") +
>   labs(x = "Fertility Rate (children per woman)", y = "Life Expectancy (years)", 
>        size = "Total Population", colour = "Region")
> ```
> 
> ```{r, echo=FALSE, message=FALSE, warning=FALSE}
> gapminder1960to2010 %>% 
>   mutate(population_total = population_male + population_female) %>% 
>   filter(year %in% c(1960, 1990, 2010)) %>% 
>   ggplot(aes(children_per_woman, life_expectancy)) +
>   geom_point(aes(size = population_total, colour = world_region)) +
>   facet_grid(cols = vars(year)) +
>   scale_colour_brewer(palette = "Dark2") +
>   labs(x = "Fertility Rate (children per woman)", y = "Life Expectancy (years)", 
>        size = "Total Population", colour = "Region")
> ```
> 
> 2. One of the countries seems to stand out from the general trend in 1960. 
> What country is it? 
> 3. Make a line plot showing how the `life_expectancy` of that country changed over 
>    time compared to another country of your choice.
> 
> > ## Answer 
> > 
> > A1. Here is the fixed code:
> > 
> > ```{r, eval=FALSE}
> > gapminder1960to2010 %>% 
> >   mutate(population_total = population_male + population_female) %>% 
> >   filter(year %in% c(1960, 1990, 2010)) %>% 
> >   ggplot(aes(children_per_woman, life_expectancy)) +
> >   geom_point(aes(size = population_total, colour = world_region)) +
> >   facet_grid(cols = vars(year)) +
> >   scale_colour_brewer(palette = "Dark2") +
> >   labs(x = "Fertility Rate (children per woman)", y = "Life Expectancy (years)", 
> >        size = "Total Population", colour = "Region")
> > ```
> > 
> > A2. Based on the x and y values in the graph, we can set a condition with `filter()`,
> > for example:
> > 
> > ```{r}
> > gapminder1960to2010 %>% 
> >   filter(year == 1960 & children_per_woman < 5 & life_expectancy < 35) %>% 
> >   select(country, year, children_per_woman, life_expectancy)
> > ```
> > 
> > A3. Here is a comparison between China and another country:
> > 
> > ```{r}
> > gapminder1960to2010 %>% 
> >   filter(country %in% c("China", "Portugal")) %>% 
> >   ggplot(aes(x = year, y = life_expectancy)) +
> >   geom_line(aes(colour = country))
> > ```
> {: .solution}
{: .challenge}

<!--
> How did the difference in life expectancy between male and female change over time for each country? 
> (hint: use the aesthetic `group = country` within `geom_line()` to have a line for each country).
> > 
> > Here is the line plot showing the change of this difference over time. 
> > There are two countries that have very abrupt change in the 80s 
> > ([possible reason](https://en.wikipedia.org/wiki/Iran%E2%80%93Iraq_War)). 
> > 
> > ```{r}
> > gapminder1960to2010 %>% 
> >   mutate(life_expectancy_female_minus_male = life_expectancy_female - life_expectancy_male) %>% 
> >   ggplot(aes(year, life_expectancy_female_minus_male)) +
> >   geom_line(aes(group = country))
> > ```
-->


## Filtering Missing Values

As noted in the callout box above, the `is.na()` function can be used to ask the 
question of whether a value is missing or not (`!is.na()`). 

For example, the following would return the rows where `main_religion` is _not_
missing.

```{r, eval=FALSE}
gapminder1960to2010 %>% 
  filter(!is.na(main_religion))
```


> ## Exercise 
> 
> - How many observations contain values for both `life_expectancy` _and_ `children_per_woman`?
> 
> > ## Answer
> > 
> > ```{r}
> > gapminder1960to2010 %>% 
> >   filter(!is.na(life_expectancy) & !is.na(children_per_woman)) %>% 
> >   # (optional) pipe the output to nrow() to get the number of rows directly
> >   nrow()
> > ```
> {: .solution}
{: .challenge}


## Using conditionals with `ifelse`

Conditions are not just useful for filtering rows, but also to modify variables or 
highlight certain observations in the data. 

For example, take the following graph showing the change in income per person:

```{r}
gapminder1960to2010 %>% 
  ggplot(aes(x = year, y = income_per_person)) +
  geom_line(aes(group = country))
```

We can highlight one of the countries by creating a new column (`mutate()`) that stores 
the result of a condition:

```{r}
gapminder1960to2010 %>% 
  # the "is_qatar" column will contain TRUE/FALSE values
  mutate(is_qatar = country == "Qatar") %>% 
  ggplot(aes(x = year, y = income_per_person)) +
  geom_line(aes(group = country, colour = is_qatar))
```

When we want to define two different outcomes from a condition, then the `ifelse()` 
function is very handy. This function takes 3 arguments: the condition we want to test, 
the values it should take if it's TRUE, and the values it should take if it's FALSE. 

For example, earlier we saw that `life_expectancy_male` has missing values encoded as 
-999. 

```{r}
summary(gapminder1960to2010$life_expectancy_male)
```

We can fix this problem using `ifelse()` inside a `mutate()` call:

```{r}
gapminder_clean <- gapminder1960to2010 %>% 
  mutate(life_expectancy_male = ifelse(life_expectancy_male == -999, NA, life_expectancy_male))

summary(gapminder_clean$life_expectancy_male)
```

In the call above we stated: if `life_expectancy_male` is equal to -999 then return 
the value `NA`, otherwise return the respective value of `life_expectancy_male` itself. 

Here is another example, where we highlight two countries in a graph by using `ifelse()`:

```{r, warning=FALSE}
gapminder1960to2010 %>% 
  mutate(income_total = population_male*income_per_person,
         country_highlight = ifelse(country %in% c("China", "United States"), country, "Others")) %>% 
  ggplot(aes(year, income_total)) +
  geom_line(aes(group = country, colour = country_highlight))
```



> ## Exercise
> 
> Using the main `dplyr` verbs that we have learned about (`select()`, `mutate()`, 
> `filter()`), create a clean version of the table called `gapminder_clean`, that 
> fulfills the following criteria:
> 
> - Does not contain the column `country_id`.
> - Contains a new column called `population_total`, which is the sum of `population_male`
>   and `population_female`.
> - All values in `main_religion` are lowercase and there are no typos. (hint:
>   use `str_*` functions to help)
> - The value -999 in `life_expectancy_male` is encoded as `NA` instead.
> - The `life_expectancy` column is numeric.
> - Remove rows with missing values for `income_groups`.
> 
> Bonus (optional):
> 
> - Make values in `main_religion` Title Case (with a space separating the words).
> - Make the `income_groups` column contain the categories: "Low", "Lower Middle", "Upper Middle", "High".
> - Change the order of the columns so that `population_total` is before the `population_male` 
>   column (the other columns should remain in the same order).
> 
> > ## Answer 
> > 
> > Here is the full pipeline to clean our data:
> > 
> > ```{r}
> > gapminder_clean <- gapminder1960to2010 %>% 
> >   select(-country_id) %>% 
> >   mutate(population_total = population_male + population_female,
> >          main_religion = str_to_lower(str_squish(main_religion)),
> >          life_expectancy_male = ifelse(life_expectancy_male == -999, NA, life_expectancy_male),
> >          life_expectancy_female = as.numeric(life_expectancy_female)) %>% 
> >   filter(!is.na(income_groups))
> > ```
> > 
> > When building such a long chain of commands, it's a good idea to check each step 
> > at a time, to see that we're doing the right thing. 
> > So, let's go through each step at a time and check the results of each operation. 
> > 
> > We used `select()` to _unselect_ the column `coutry_id` (using the `-` sign 
> > before the column name):
> > 
> > ```{r}
> > gapminder1960to2010 %>% 
> >   select(-country)
> > ```
> > 
> > We used mutate to calculate `population_total`. To confirm we're doing the right thing, 
> > we could use the function `select()` to check that we're getting the expected result:
> > 
> > ```{r}
> > gapminder1960to2010 %>% 
> >   mutate(population_total = population_male + population_female) %>% 
> >   select(population_male, population_female, population_total)
> > ```
> > 
> > To fix the values in `main_religion` we used functions from the `stringr` package. 
> > We can use the function `distinct()` to check our results:
> > 
> > ```{r}
> > gapminder1960to2010 %>% 
> >   mutate(main_religion = str_to_lower(str_squish(main_religion))) %>% 
> >   distinct(main_religion)
> > ```
> > 
> > To fix `life_expectancy_male` we used an `ifelse()`. We can check that we're 
> > getting the right result by using `arrange()` to sort our table and check that we 
> > no longer have values of -999:
> > 
> > ```{r}
> > gapminder1960to2010 %>% 
> >   mutate(life_expectancy_male = ifelse(life_expectancy_male == -999, NA, life_expectancy_male)) %>% 
> >   arrange(life_expectancy_male) %>% 
> >   select(life_expectancy, life_expectancy_male)
> > ```
> > 
> > Finally, we can use the `as.numeric()` function to convert `life_expectancy_female` 
> > to numbers. We get a warning  because the function introduces `NA` when it finds 
> > the value "-" (which is impossible to be converted, so it is encoded as missing).
> > Again, we can use the `arrange()` function to check that we no longer get the value 
> > "-" (try running `arrange()` with and without the `mutate()` step).
> > 
> > ```{r}
> > gapminder1960to2010 %>% 
> >   mutate(life_expectancy_female = as.numeric(life_expectancy_female)) %>% 
> >   arrange(life_expectancy_female) %>% 
> >   select(life_expectancy, life_expectancy_female)
> > ```
> > 
> > After checking all of these individual steps work correctly, we can then assemble 
> > them together into a chain, as we did above.
> {: .solution}
{: .challenge}

## Saving data

Now that we have a clean version of our table, it's a good idea to save it for 
future use. You can use the `write_*()` family of functions to save data in a variety 
of formats. 

Let's use `write_csv()` as an example. 
The `write_csv()` function needs the name of the table you want to save and then 
_path_ to the file you want to save it in (don't forget the file extension!):

```{r, eval=FALSE}
write_csv(gapminder_clean, "data/processed/gapminder1960to2010_clean.csv")
```

There are many other functions for saving data, you can check the documentation 
with `?write_delim`. 


> ## Data Tip: Cleaning Data
> 
> The infamous [80/20 rule](https://whatsthebigdata.com/2016/05/01/data-scientists-spend-most-of-their-time-cleaning-data/) 
> in data science suggests that about 80% of the time is spend preparing the data
> for analysis. While this is not really a scientific rule, it does have some relation 
> to the real life experience of data analysts. 
> 
> Although it's a lot of effort, and usually not so much fun, if you make sure to clean
> and format your data correctly, it will make your downstream analysis much more fluid,
> fruitful and pleasant.
> 
{: .discussion}
