---
title: "Apendix: multidimensional data analysis"
teaching: 50
exercises: 30
questions:
  - "How to perform Principal Components Analysis (PCA)?"
  - "How to perform hierarchical clustering?"
  - "How to perform k-means clustering?"
objectives:
  - "Recognise when to use grouped operations in data analysis."
  - "Differentiate between grouped summaries and other types of grouped operations."
  - "Apply grouped summaries using the `group_by()` + `summarise()` functions."
  - "Apply other grouped operations such as: `group_by()` + `filter()` and `group_by()` + `mutate()`."
  - "Recognise the importance of the `ungroup()` function."
keypoints:
  - ""
source: Rmd
---

  ```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("09-")
knitr::opts_chunk$set(fig.width=12, fig.height=7)
```

```{r, include=FALSE}
# set ggplot2 theme with bigger text size
library(tidyverse)
theme_set(theme_grey() + theme(text = element_text(size = 16)))
```


In this lesson we're going to learn how to use the `dplyr` package to make calculations
for sub-groups in our data.

As usual when starting an analysis on a new script, let's start by loading the
packages and reading the data. In this case, let's use the clean dataset that we
created in the last exercise of the
[previous episode]({{ page.root }}{% link _episodes/05-manipulate_observations_dplyr.md %}).

```{r, eval=FALSE}
# load the package
library(tidyverse)

# Read the data, specifying how missing values are encoded
gapminder_clean <- read_csv("data/processed/gapminder1960to2010_socioeconomic_clean.csv",
                            na = "")
```

If you haven't completed the previous exercise, here's how you can recreate the
clean dataset:

```{r, message=FALSE, warning=FALSE}
gapminder_clean <- read_csv("data/gapminder1960to2010_socioeconomic.csv", na = "") %>%
  select(-country_id) %>%
  mutate(population_total = population_male + population_female,
         main_religion = str_to_lower(str_squish(main_religion)),
         life_expectancy_male = ifelse(life_expectancy_male == -999, NA, life_expectancy_male),
         life_expectancy_female = as.numeric(life_expectancy_female)) %>%
  filter(!is.na(income_groups))
```


## PCA to create an alternative HDI

```{r}
pca <- gapminder_clean %>%
  filter(year == 2009) %>%
  select(country, income_per_person, life_expectancy, school_years_men, school_years_women) %>%
  # pivot_wider(names_from = "year",
  #             values_from = c("income_per_person", "life_expectancy", "school_years_women", "school_years_men")) %>%
  column_to_rownames("country") %>%
  drop_na() %>%
  prcomp(scale = TRUE)

library(ggfortify)
autoplot(pca, loadings = TRUE, loadings.label = TRUE)

library(broom)

# makes sense that PC4 hardly explains anything, given school years male/female being very highly correlated
tidy(pca, matrix = "pcs") %>%
  ggplot(aes(PC, percent)) +
  geom_col() +
  geom_line(aes(y = cumulative))

# We can see how this creates an alternative score for HDI
pca$x %>%
  as_tibble(rownames = "country") %>%
  inner_join(gapminder_clean, by = c("country")) %>%
  filter(year == 2005) %>%
  #with(cor(hdi_human_development_index, PC1, use = "complete.obs", method = "spearman"))
  ggplot(aes(PC1, hdi_human_development_index)) +
  geom_point()
```


## Hierarchichal clustering to replace income groups

```{r}
clust <- gapminder_clean %>%
  filter(year > 2000) %>%
  select(country, year, income_per_person) %>%
  pivot_wider(names_from = "year",
              names_prefix = "yr",
              values_from = "income_per_person") %>%
  column_to_rownames("country") %>%
  as.matrix() %>%
  log() %>%
  dist() %>%
  hclust("ward.D2")

plot(clust)
rect.hclust(clust, k = 4)
```

```{r}
cutree(clust, k = 4) %>% enframe(name = "country", value = "cluster") %>%
  full_join(gapminder_clean, by = "country") %>%
  ggplot(aes(cluster, income_groups)) +
  geom_count()
```

```{r}
cutree(clust, k = 4) %>% enframe(name = "country", value = "cluster") %>%
  full_join(gapminder_clean, by = "country") %>%
  ggplot(aes(year, income_per_person)) +
  geom_line(aes(group = country, colour = factor(cluster))) +
  scale_y_log10()
```


## PCA per year!

```{r}
# function to do the PCA
pca_gapminder <- function(x){
  pca <- x %>%
    column_to_rownames("country") %>%
    drop_na()

  if (nrow(pca) == 0){
    return(data.frame(PC1 = NA, PC2 = NA, PC3 = NA, PC4 = NA))
  } else {
    pca <- pca %>%
      prcomp(scale = TRUE)
    return(as_tibble(pca$x, rownames = "country"))
  }
}

gapminder_clean %>%
  select(country, year, income_per_person, life_expectancy,
         school_years_men, school_years_women) %>%
  drop_na() %>%
  group_by(year) %>%
  nest() %>%
  mutate(pca_out = map(data, pca_gapminder)) %>%
  select(-data) %>%
  unnest(pca_out) %>%
  ggplot(aes(year, PC1)) +
  geom_line(aes(group = country, colour = country == "United Arab Emirates"))
```



