---
title: "Extra practice exercises"
teaching: 00
exercises: 00
questions:
  - "How to apply the tools and concepts learned to new data?"
objectives: 
  - "Use R and the `tidyverse` collection of packages to explore a new dataset."
keypoints:
  - "The initial exploration of data is crucial to detect any data quality issues that need fixing."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("90-")
```

<br>

This document suggests new exercises to practice the skills and concepts learned 
in the course. 
Each exercise section links to one or several of the main lesson episodes it refers to.

(**Note:** solutions are not yet available for these exercises and will be added in time)


## Starting a data analysis project

This exercise uses skills learned in the episodes 
[Introduction to R and RStudio]({{ page.root }}{% link _episodes/00-starting_your_project.md %}),
[Basic objects and data types in R]({{ page.root }}{% link _episodes/01-objects_functions_vectors.md %}) 
and [Working with Tabular Data]({{ page.root }}{% link _episodes/02-data_frames.md %}).

For this set of questions, we will use a new dataset from the Gapminder Foundation, 
which includes data about energy consumption and CO2 emmissions in the different 
countries. 
This new data file includes the following variables:

| Column                       | Description                             |
|------------------------------|-----------------------------------------|
| country                      | country name                            |
| world_region                 | 6 world regions                         |
| year                         | year that each datapoint refers to      |
| yearly_co2_emissions         | yearly CO2 emissions, 1000 metric tones |
| energy_use_per_person        | energy use per person                   |
| energy_production_per_person | energy production per person            |


### Part I - setting up your project

1. On your computer, create a new directory for this project, for example `gapminder-energy`
2. Create a new R Project on this directory
3. Create directories that you feel are important, including one for saving the raw data.
4. Download the [gapminder energy dataset](https://raw.githubusercontent.com/tavareshugo/r-eda-gapminder/gh-pages/_episodes_rmd/data/raw/gapminder1990to2010_energy.tsv) 
and save it in your project's data folder. 
(<kbd>right-click</kbd> the file link and choose "Save link as...")
5. Install the `skimr` package, which you will use to do some quality control checks 
on the data file. (Hint: `install.packages()`)


### Part II - reading and checking quality of data

1. Read the `gapminder1990to2010_energy.tsv` file into a `data.frame`/`tibble` object 
called `energy`. (Hint: note that this is a _tab-delimited_ file.)
2. Make a basic check that your data import went well. Your data frame should have 
3968 rows and 7 columns. (Hint: use `summary()` and `str()` to check data integrity)
3. Use the `skim()` function from the `skimr` package to get an informative summary
of your data frame. Make sure columns were imported as the correct type of value, 
and that you have the expected number of categories for character columns. 
(Hint: don't forget to load the package first with `library()`)

Note any problems you find during this exploratory analysis.


## Data Visualisation

Following from the previous exercise on the `gapminder1990to2010_energy.tsv` dataset:

1. Use a histogram to investigate the distribution of yearly CO2 emissions in the dataset. 
Consider whether the original scale is adequate. 
2. Has the distribution of yearly CO2 emissions changed dramatically over the years? 
Try using `geom_density()` or `geom_freqpoly()` with the following aesthetic `group = year`.
(to ensure one line is drawn for each year).
3. Look at the trend over time (x-axis) of CO2 emissions (y-axis) for each country, 
both on a linear and on a log scales. How does the scale change your perception of the data? 
(hint use `geom_line()` with `aes(group = country)`).
4. Try and recreate the following graph, showing the relationship between energy use 
and CO2 emissions. Note that the axis are on a log-scale. The "Dark2" palette was used 
from the _brewer_ colour scale. 

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=15, fig.height=15}
library(tidyverse)
energy <- read_tsv("data/raw/gapminder1990to2010_energy.tsv", na = "")

energy %>% 
  ggplot(aes(energy_use_per_person, yearly_co2_emissions)) +
  geom_point(aes(colour = world_region)) +
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") +
  scale_colour_brewer(palette = "Dark2") +
  facet_wrap(facets = vars(year)) +
  labs(x = "Energy use per Person", y = "Total CO2 emissions (tons)",
       colour = "World Region", tag = "A") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 20))
```


## Data manipulation 

The `gapminder1990to2010_energy.tsv` dataset contains some mistakes that need correcting. 

Create a new table, `energy_clean`, that has the following issues fixed:

- Convert the `energy_production_per_person` from character to numeric. 
  (hint: use the `as.numeric()` function within `mutate()`)
- `country` has 191 unique categories, whereas `country_id` has 190 (you can check 
  this with `unique(energy$country)` and `unique(energy$country_id)`). One of the 
  countries, _Brazil_ is spelled with an "s" and a "z". Change the values of "Brasil" 
  to "Brazil". (hint: use the `ifelse()` function within `mutate()`)

From the new table, try and fix the following code, by writing the correct condition 
in `filter()`, which produces the plot below:

```{r, eval=FALSE}
energy_clean %>% 
  filter(FIXME) %>% 
  ggplot(aes(energy_use_per_person, energy_production_per_person)) +
  geom_point(aes(colour = world_region)) + 
  facet_wrap(facets = vars(year)) +
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") +
  scale_colour_brewer(palette = "Dark2")
```

```{r, echo=FALSE, warning=FALSE}
energy_clean <- energy %>% 
  mutate(energy_production_per_person = as.numeric(energy_production_per_person),
         country = ifelse(country == "Brasil", "Brazil", country))

energy_clean %>% 
  filter(year <= 1992 | year >= 2008) %>% 
  ggplot(aes(energy_use_per_person, energy_production_per_person)) +
  geom_point(aes(colour = world_region)) + 
  facet_wrap(facets = vars(year)) +
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") +
  scale_colour_brewer(palette = "Dark2")
```


{% include links.md %}
